The Application of Recurrent Quantification Analysis to Human-Humanoid Interaction

### 2019-04-29 docs/manuscript/main.tex
[docs/corrections/29-03-2019] part-01

\title{Strengths and Weaknesses of Recurrent Quantification Analysis 
in the context of Human-Humanoid Interaction}
\author[1,*]{Miguel Xochicale}
\author[1]{Chirs Baber}
\affil[1]{University of Birmingham, 
	School of Engineering, 
	Birmingham, 
	B15 2TT, 
	UK}
\affil[*]{map479@bham.ac.uk}

@@

\begin{abstract}
Human movement variability occurs in motor performance 
across multiple repetitions of a task and such behaviour is an
inherent feature within and between each persons' movement. Quantifying
movement variability is still an open problem, particularly when traditional
methods in time domain and frequency domain fail to detect tiny modulations in
frequency or phase for time series. 
For this work, we hence investigate methodologies from nonlinear dynamics such as 
reconstructed state space (RSS), uniform time-delay embedding (UTDE), 
recurrence plots (RPs) and recurrence quantification analysis (RQA) metrics. 
Particularly, we are interested in the weaknesses and robustness 
of nonlinear dynamics tools from raw and post-processed data of 
wearable inertial sensors (IMUs). 
In the reported experiment, twenty right-handed healthy participants 
imitated simple vertical and horizontal arm movements in normal 
and faster speed from an humanoid robot.
With four window lengths and three levels of smoothed time series, 
we found visual differences in the patterns of RSSs and RPs. 
Then using metrics of RQA, we find out that the type of movements
and the level of smoothness affects those metrics. 
Our contribution for this work is that entropy values from RQA were 
well distributed and presenting variation in all the conditions for time series. 
This work have then the potential to enhance the development of 
better diagnostic tools for various applications rehabilitation, 
sport science or for new forms of human-robot interaction.
\end{abstract}
%226

@@

\section*{Introduction}
Human movement variability is a complex system where not only multiple
joints and limbs are involved for a specific task in a determined environment
but also how we process the external information with all of our available senses 
and use our experiences \cite{davids2003}. 
Recent studies in human motion recognition have revealed the possibility to 
estimate features from lower dimensions signals to distinguish differences between 
styles of pedalling motion \cite{Quintana-Duque2012, Quintana-Duque2016}, 
gait identification \cite{sama2013, frank2010} 
or pattern recognition from biological signals \cite{gomezgarcia2014}.
The lower dimension signals from biological signals are time series 
of $1-$dimension in $\mathbb{R}$ which commonly are 
noisy, nonlinear and non-stationary \cite{gomezgarcia2014}.
Hence, we are hypothesising that nonlinear dynamics can be used to 
objectively quantify such human movement variability of lower
dimension signals \cite{Quintana-Duque2012, Quintana-Duque2016, sama2013, 
frank2010, gomezgarcia2014, marwan2011, stergiou2011}.
For instance, Bradley et al. 2015 \cite{bradley2015} reviewed methods for
nonlinear time series analysis based on the appropriate estimation of the 
embedding parameters ($m$ embedding dimension and $\tau$ embedding delay) to 
reconstruct the state space,
where an $n$-dimensional reconstructed state space using $1-$dimensional 
time series,
can preserve the topological properties of an unknown $M$-dimensional 
state space \cite{takens1981}.
Also, Bradley et al. 2015 \cite{bradley2015} reviewed the use of 
Recurrence Plots (RPs), a graphical representation of a two-dimensional map 
which show black and white dots as recurrences in a given $n$-dimensional system, 
and Recurrence Quantification Analysis (RQA) metrics can pick out important 
directions and statistics in RPs.
In general, RPs and RQA provide a more intuitive meaning of the time series,
for instance, RQA is quantitatively and qualitatively independent of 
embedding dimension (also verified experimentally \cite{iwanski1998}).
However, the estimation of embedding parameters and the selection of the 
right parameters to perform RQA is still an open problem.
Bradley et al. 2015 \cite{bradley2015} pointed out that
there is no general technique that can be used to compute the embedding
parameters since the time series are system-dependent which means that
once computing the values for embedding parameters, these may only work for
one purpose (e.g., prediction) and may not work well for another purpose
(e.g., computing dynamical invariants).
Additionally, the methodologies of nonlinear dynamics for 
computing the embedding parameters e.g., autocorrelation, mutual information, 
and nearest neighbour require data which is well 
sampled and with little noise \cite{garland2016}
or require purely deterministic signals \cite{kantz2003}.
Similarly, these nonlinear analysis methods for computing the embedding 
parameters can break down
with real-world datasets which have generally different length, 
different values of accuracy and precision (rounding errors due to finite 
precision of the measurement apparatus which include frequency 
acquisition \cite{frank2010}),
and data may be contaminated with different sources of noise
\cite{garland2016}.
It is surprising that even with the previous constraints with regard to 
the quality of data, and the problem with the estimation of embedding parameters,
the use of nonlinear dynamics have proven to be helpful to understand and 
characterise dynamics of time series 
\cite{Quintana-Duque2012, Quintana-Duque2016, sama2013, frank2010,
gomezgarcia2014, gomezgarcia2014, marwan2011, stergiou2011, bradley2015}.

@@

%*******************************************************************************
%*******************************************************************************
%*******************************************************************************
\section*{Results}
To give insight to the research questions, 
we considered various conditions of time series collected for this work 
(see Experiment section for more details) which are described as follows
\begin{itemize}
\item Three levels of smoothness for the normalised data 
	(sg0zmuv, sg1zmuv and sg2zmuv), computed from two different filter 
	lengths (29 and 159) with the same polynomial degree 
	of 5 using the function \texttt{sgolay(p,n,m)} \cite{Rsignal},
\item four velocities arm movement activities: 
	horizontal normal (HN), horizontal faster (HF), 
	vertical normal (VN), and vertical faster (VF), and
\item four window lengths: 2-sec (100 samples), 5-sec (250 samples), 
	10-sec (500 samples) and 15-sec (750 samples).
\end{itemize}

%\subsection*{Time series}
After the data collection, raw time series were windowed, normalised and 
smoothed. Then, due to space limitations and to have simple visualisation, 
we only present 10-sec (500 samples) window length time series for 
three participants (p01, p01 and p03) performing horizontal 
arm movements (axis GyroZ) and vertical arm movements (axis GyroY) (Figs \ref{fig:ts}). 
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/timeseries/pdf/fig1}
    	\caption{ 
	{\bf Time series for horizontal and vertical arm movements.}
		(A/D) raw-normalised (sg0zmuv), 
		(B/E) normalised-smoothed 1 (sg1zmuv) and
		(C/F) normalised-smoothed 2 (sg2zmuv).
		Time series are only for three participants 
		($p01$, $p02$, and $p03$) 
		for horizontal and vertical arm movements in normal 
		and faster velocity (HN, HF, VN, VF) 
		with the normalised GyroZ or GyroY axis 
		and with one sensor attached to the participant (HS01) 
		and other sensor attached to the robot (RS01).	
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:ts}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

%\subsection*{Reconstructed State Spaces using Uniform Time-Delay Embedding}
\subsection*{Reconstructed State Spaces}
One of the main challenges of the implementation of 
Uniform Time-Delay Embedding to Reconstructed State Spaces is the 
selection of embedding parameters because each time series is unique in 
terms of its structure (modulation of amplitude, frequency and phase) 
\cite{ frank2010, sama2013, bradley2015}.
With that in mind, the problem for this work is not to compute individual 
embedding parameters for each of the time series but to deal with a 
selecting of two parameters that can represent all the time series. 
Our solution for that problem was to compute a sample mean over all 
values in each of the conditions of the time series 
%of Figs~\ref{fig:ts}
for minimum dimension values and 
%Figs~\ref{fig:caoami} 
for minimum delay values as follows.
%, resulting in an average minimum embedding parameters 
%of ($\overline{m_0}=6$, $\overline{\tau_0}=8$).

\subsubsection*{Minimum embedding parameters}
Minimum embedding parameters were firstly computed 
with False Nearest Neighbour and Average Mutual Information.
%for Uniform Time-Delay Embedding (UTDE),
%Reconstructed State Space (RSS), 
%Recurrence Plots (RPs) and 
%Recurrence Quantification Analysis (RQAs).
Figures \ref{fig:caoami}(A) show that minimum embedding values 
for sensor HS01 appear to show more variations as their 
interquartile range of the box plots are near to 1 
with three exceptions (HF sg0, VN sg0, and VF sg1).
Minimum embedding values appear to be constant for sensor RS01 
as their interquartile range 
in the box plots are near to 0.1 with the exception of 
two axis (HF sg0 and VF sg1).
Additionally, it can be seen in Figs \ref{fig:caoami}(A) a 
decrease of mean values (rhombus) in the box plots
as smoothness of time series increase.
Similarly, the first minimum values of AMI values are 
shown in the box plots of Figs \ref{fig:caoami}(B).
It can be seen that values for HS01 tend to be more spread as the smoothness 
of the time series is increasing 
(see the increase of both mean (rhombus) and interquartile range).
However, AMI values for RS01 do not show such increase in relation with
the increase of smoothness excepting for HF and VF
(Figs \ref{fig:caoami}(B)).

@@

Evidently, it is easy to observe by eye the differences in each of the
trajectories in the reconstructed state spaces 
(Figs~\ref{fig:rss_aHw10}, \ref{fig:rss_aVw10}), 
however one might be not objective when quantifying those differences 
since those observation might vary from person to person.
With that in mind, we tried to objectively quantify those differences 
using euclidean distances between the origin to each of the points in the 
trajectories, however these results created a suspicious metric, specially 
for trajectories which looked very messy.
With that in mind, we computed Recurrence Plots and 
Recurrence Quantification Analysis to objectively quantify 
the differences in each of the cases of the time series.

@@

\subsection*{Other effects on 3D RQA ENTR values}
% metrics with different embedding parameters, 
%	recurrence thresholds, window lengths, levels of smoothness, 
%	and time series structures.}

@@

%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/rqas}
%%\includegraphics[width=1.0\textwidth]{rqas}
%    \caption{
%	{\bf 3D surfaces for RQA metrics.}
%	3D surfaces for REC, DET, RATIO and ENTR values with increasing 
%	pair embedding parameters ($0 \ge m \le 10$, $0 \ge \tau \le 10$) 
%	and recurrence thresholds (  $ 0.2 \ge \epsilon \le 3 $).
%	RQA metrics values for time series of participant p01 using HS01 sensor, 
%	HN activity and sg0zmuvGyroZ axis and 500 samples window length.
%        R code to reproduce the figure is available from \cite{hwum2018}.
%	}
%\label{fig:topo_rqas}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------

%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/w}
%%\includegraphics[width=1.0\textwidth]{w}
%    \caption{
%	{\bf 3D surfaces for RQAs metrics with four window lengths.}
%	3D surfaces of RQAs metric values with increasing embedding 
%	parameters and recurrence thresholds are for four window lengths 
%	(w100, w250, w500 and  w750).
%	RQA metrics values are for time series of participant p01 
%	using HS01 sensor, HN activity and sg0zmuvGyroZ axis.
%	R code to reproduce the figure is available from \cite{hwum2018}.
%        }
%\label{fig:topo_sensoractivities}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/s}
%%\includegraphics[width=1.0\textwidth]{s}
%    \caption{
%	{\bf 3D surfaces for RQA metrics with three levels of smoothness.}
%       	3D surfaces of RQAs metric values with increasing embedding 
%	parameters and recurrence thresholds are for three levels of 
%	smoothness (sg0zmuvGyroZ, sg1zmuvGyroZ and sg1zmuvGyroZ).
%	RQA metrics values are for time series of participant p01 using 
%	HS01 sensor, HN activity and 500 samples window length.
%	R code to reproduce the figure is available from \cite{hwum2018}.
% }
%\label{fig:topo_smoothness}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------
%
%
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/results/rqa/v03/rqa_plots/topologies/figs/p}
%%\includegraphics[width=1.0\textwidth]{p}
%    \caption{
%	{\bf 3D surfaces for RQA metrics with three participants.}
%       	3D surfaces of RQAs metric values for participants p01, p02 and p03 
%	with increasing embedding parameters and recurrence thresholds.
%	RQA metrics values are for time series of HS01 sensor, 
%	HN activity and 500 samples window length.
%	R code to reproduce the figure is available from \cite{hwum2018}.
% }
%\label{fig:topo_participants}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------
%
%
%
%
%
%
%
%
%
%
%
%*******************************************************************************
%*******************************************************************************
%*******************************************************************************
\section*{Discussion}
%The Discussion should be succinct and must not contain subheadings.

@@

It is evidently that time series from different sources 
(participants, movements, axis type, window length or levels of smoothness) 
presents visual differences for embedding parameters and therefore for RRSs. 
For which, the selection of embedding parameters was our first challenge 
where we computed embedding parameters for each time series 
(Fig \label{fig:caoami}) and then computed a sample mean over 
all time series in order to get two embedding parameters 
to compute all RRSs (Figs \label{fig:rss_aHw10} and \label{fig:rss_aVw10}). 
%with its corresponded type of movement. 
Then we found that the quantification of variability with regard to 
the shape of the trajectories in RSSs requires more investigation 
since our original proposed method base on euclidean metric failed 
to quantify those trajectories. Specially, for trajectories which 
were not well unfolded. 
With that in mind, we proceed to apply 
Recurrence Quantification Analysis (RQA) metrics 
(REC, DET, RATIO and ENTR) in order to avoid any subjective interpretations 
or personal bias with regard to the representation of the trajectories in RSSs.

@@

\subsection*{Data from Inertial Measurement Units} 
	\label{sec:experiment:subsec:imu}
\subsubsection*{Raw data}
Considering the work of \cite{shoaib2016} which provided evidence of an 
improvement in recognition activities when combining data from 
accelerometer and gyroscope. We focus our analysis from data of the 
accelerometer and gyroscope of the NeMEMsi sensors \cite{Comotti2014} and 
leave the data of the magnetometer and quaternions for further investigation 
because of their possible variations with regard to magnetic disturbances.

>>>>

\title{The Application of Recurrent Quantification Analysis 
to Human-Humanoid Interaction}
\author[1,*]{Miguel Xochicale}
\author[1]{Chris Baber}
\affil[1]{University of Birmingham, 
	School of Engineering, 
	Birmingham, 
	B15 2TT, 
	UK}
\affil[*]{map479@bham.ac.uk}

@@

\begin{abstract}
Human movement variability occurs in motor performance 
across multiple repetitions of a task and such behaviour is an
inherent feature within and between each persons' movement. Quantifying
movement variability is still an open problem, particularly when traditional
methods in time domain and frequency domain fail to detect tiny modulations in
frequency or phase for time series. 
For this work, we hence investigate methodologies from nonlinear dynamics such as 
reconstructed state space (RSS), uniform time-delay embedding (UTDE), 
recurrence plots (RPs) and recurrence quantification analysis (RQA) metrics. 
Particularly, we are interested in the weaknesses and robustness 
of nonlinear dynamics tools from raw and post-processed data of 
wearable inertial sensors (IMUs). 
In the reported experiment, twenty right-handed healthy participants 
imitated simple vertical and horizontal arm movements in normal 
and faster speed from an humanoid robot.
With four window lengths and three levels of smoothed time series, 
we found visual differences in the patterns of RSSs and RPs. 
Then with the use of RQA metrics, we find out that the type of 
movements and the level of smoothness affects those metrics. 
Specifically, entropy values from RQA were well distributed and 
presented variation in all the conditions for time series. 
This work have then the potential to enhance the development of 
better diagnostic tools for various applications in rehabilitation, 
sport science or for new forms of human-robot interaction.
\end{abstract}
%226

@@

\section*{Introduction}
Human movement requires a complex system where not only multiple
joints and limbs are involved for a specific task in a determined environment
but also perception and action of movement affects the physical performance 
\cite{davids2003}. 
In contrast, variability in humanoid movement is usually very small,
as a result of the control system used. This means that, while humanoids 
solve the degrees of freedom problem through join design and algorithms,
humans tend to have more fluid and flexible approach. Consequently, 
one can see much variability in the performance of even the simplest task.
Studies of human motion reveal the possibility to 
estimate features from lower dimension signals to distinguish differences between 
styles of pedalling motion \cite{Quintana-Duque2012, Quintana-Duque2016}, 
gait identification \cite{sama2013, frank2010},
and detection of pathologies \cite{gomezgarcia2014}.
The lower dimension signals from biological systems are time series 
of $1-$dimension in $\mathbb{R}$ which commonly are 
noisy, nonlinear and non-stationary \cite{gomezgarcia2014}.
Hence, nonlinear dynamics can be used to 
objectively quantify variability of lower
dimension signals \cite{Quintana-Duque2012, Quintana-Duque2016, sama2013, 
frank2010, gomezgarcia2014, marwan2011, stergiou2011}.
For instance, Bradley et al. 2015 \cite{bradley2015} reviewed methods for
nonlinear time series analysis based on the appropriate estimation of the 
embedding parameters ($m$ embedding dimension and $\tau$ embedding delay) to 
reconstruct the state space,
where an $n$-dimensional reconstructed state space using $1-$dimensional 
time series,
can preserve the topological properties of an unknown $M$-dimensional 
state space \cite{takens1981}.
Bradley et al. 2015 \cite{bradley2015} also reviewed the use of 
Recurrence Plots (RPs), a graphical representation of a two-dimensional map 
which show black and white dots as recurrences in a given $n$-dimensional system, 
and Recurrence Quantification Analysis (RQA) metrics compute statistics in RPs.
In general, RPs and RQA provide an intuitive meaning of the time series,
for instance, RQA is quantitatively and qualitatively independent of 
embedding dimension (also verified experimentally \cite{iwanski1998}).
However, the estimation of embedding parameters and the selection of the 
right parameters to perform RQA is still an open problem.
Bradley et al. 2015 \cite{bradley2015} pointed out that
there is no general technique that can be used to compute the embedding
parameters since time series are system-dependent which means that
computing the values for embedding parameters may only work for
one purpose (e.g., prediction) and may not work well for another purpose
(e.g., computing dynamical invariants).
Additionally, the methodologies of nonlinear dynamics for 
computing the embedding parameters e.g., autocorrelation, mutual information, 
and nearest neighbour require data which is well 
sampled and with little noise \cite{garland2016}
or require purely deterministic signals \cite{kantz2003}.
Thus, these nonlinear analysis methods for computing the embedding 
parameters can break down
with real-world datasets which have generally different length, 
different values of accuracy and precision (rounding errors due to finite 
precision of the measurement apparatus which include frequency 
acquisition \cite{frank2010}),
and data may be contaminated with different sources of noise
\cite{garland2016}.
It is surprising that even with the previous constraints with regard to 
the quality of data, and the problem with the estimation of embedding parameters,
the use of nonlinear dynamics have proven to be helpful to understand and 
characterise dynamics of time series 
\cite{Quintana-Duque2012, Quintana-Duque2016, sama2013, frank2010,
gomezgarcia2014, gomezgarcia2014, marwan2011, stergiou2011, bradley2015}.

@@

%*******************************************************************************
%*******************************************************************************
%*******************************************************************************
\section*{Results}

\subsection*{Reconstructed State Spaces}
One of the challenges of the implementation of 
Uniform Time-Delay Embedding to Reconstructed State Spaces is the 
selection of embedding parameters because each time series is unique in 
terms of its structure (modulation of amplitude, frequency and phase) 
\cite{ frank2010, sama2013, bradley2015}.
With that in mind, the problem for this work is not to compute individual 
embedding parameters for each of the time series but to deal with a 
selecting of two parameters that can represent all the time series. 
Our solution for that problem was to compute a sample mean over all 
values in each of the conditions of the time series 
%of Figs~\ref{fig:ts}
for minimum dimension values and 
%Figs~\ref{fig:caoami} 
for minimum delay values as follows.
%, resulting in an average minimum embedding parameters 
%of ($\overline{m_0}=6$, $\overline{\tau_0}=8$).

\subsubsection*{Minimum embedding parameters}
Minimum embedding parameters were firstly computed 
with False Nearest Neighbour and Average Mutual Information.
%for Uniform Time-Delay Embedding (UTDE),
%Reconstructed State Space (RSS), 
%Recurrence Plots (RPs) and 
%Recurrence Quantification Analysis (RQAs).
Figures \ref{fig:caoami}(A) show that minimum embedding values 
for sensor HS01 appear to show more variations as 
interquartile range is near to 1 
with three exceptions (HF sg0, VN sg0, and VF sg1).
Minimum embedding values appear to be constant for sensor RS01 
as interquartile range is near to 0.1 with the exception of 
two axis (HF sg0 and VF sg1).
Additionally, it can be seen in Figs \ref{fig:caoami}(A) a 
decrease of mean values (rhombus) in the box plots
as smoothness of time series increase.
Similarly, the first minimum values of AMI values are 
shown in the box plots of Figs \ref{fig:caoami}(B).
It can be seen that values for HS01 tend to be more spread as the smoothness 
of the time series is increasing 
(see the increase of both mean (rhombus) and interquartile range).
However, AMI values for RS01 do not show such increase in relation with
the increase of smoothness except for HF and VF
(Figs \ref{fig:caoami}(B)).

@@

It is easy to observe by eye the differences in each of the
trajectories in the reconstructed state spaces 
(Figs~\ref{fig:rss_aHw10}, \ref{fig:rss_aVw10}), 
however one might be not objective when quantifying those differences 
since those observation might vary from person to person.
With that in mind, we tried to objectively quantify those differences 
using euclidean distances between the origin to each of the points in the 
trajectories, however these results created a suspicious metric, specially 
for trajectories which looked very messy.
With that in mind, we computed Recurrence Plots and 
Recurrence Quantification Analysis to objectively quantify 
the differences in each of the cases of the time series.

@@

\subsection*{Other effects on 3D RQA ENTR values}

@@

%*******************************************************************************
%*******************************************************************************
%*******************************************************************************
\section*{Discussion}
%The Discussion should be succinct and must not contain subheadings.

@@

It is evident that time series from different sources 
(participants, movements, axis type, window length or levels of smoothness) 
presents visual differences for embedding parameters and therefore for RRSs. 
For which, the selection of embedding parameters was our first challenge 
where we computed embedding parameters for each time series 
(Fig \label{fig:caoami}) and then computed a sample mean over 
all time series in order to get two embedding parameters 
to compute all RRSs (Figs \label{fig:rss_aHw10} and \label{fig:rss_aVw10}). 
%with its corresponded type of movement. 
Then we found that the quantification of variability with regard to 
the shape of the trajectories in RSSs requires more investigation 
since our original proposed method base on euclidean metric failed 
to quantify those trajectories. Specially, for trajectories which 
were not well unfolded. 
With that in mind, we apply 
Recurrence Quantification Analysis (RQA) metrics 
(REC, DET, RATIO and ENTR) in order to avoid any subjective interpretations 
or personal bias with regard to the representation of the trajectories in RSSs.

@@

\subsection*{Data from Inertial Measurement Units} 
	\label{sec:experiment:subsec:imu}

To give insight to the research questions, 
we considered various conditions of time series collected for this work 
(see Experiment section for more details) which are described as follows
\begin{itemize}
\item Three levels of smoothness for the normalised data 
	(sg0zmuv, sg1zmuv and sg2zmuv), computed from two different filter 
	lengths (29 and 159) with the same polynomial degree 
	of 5 using the function \texttt{sgolay(p,n,m)} \cite{Rsignal},
\item four velocities arm movement activities: 
	horizontal normal (HN), horizontal faster (HF), 
	vertical normal (VN), and vertical faster (VF), and
\item four window lengths: 2-sec (100 samples), 5-sec (250 samples), 
	10-sec (500 samples) and 15-sec (750 samples).
\end{itemize}
%\subsection*{Time series}
After the data collection, raw time series were windowed, normalised and 
smoothed. Then, due to space limitations and to have simple visualisation, 
we only present 10-sec (500 samples) window length time series for 
three participants (p01, p01 and p03) performing horizontal 
arm movements (axis GyroZ) and vertical arm movements (axis GyroY) (Figs \ref{fig:ts}). 
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/timeseries/pdf/fig1}
    	\caption{ 
	{\bf Time series for horizontal and vertical arm movements.}
		(A/D) raw-normalised (sg0zmuv), 
		(B/E) normalised-smoothed 1 (sg1zmuv) and
		(C/F) normalised-smoothed 2 (sg2zmuv).
		Time series are only for three participants 
		($p01$, $p02$, and $p03$) 
		for horizontal and vertical arm movements in normal 
		and faster velocity (HN, HF, VN, VF) 
		with the normalised GyroZ or GyroY axis 
		and with one sensor attached to the participant (HS01) 
		and other sensor attached to the robot (RS01).	
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:ts}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsubsection*{Raw data}
Considering the work of \cite{shoaib2016} which provided evidence of an 
improvement in recognition activities when combining data from 
accelerometer and gyroscope. We focus our analysis from data of the 
accelerometer and gyroscope of the NeMEMsi sensors \cite{Comotti2014} and 
leave the data of the magnetometer and quaternions for further investigation 
because of their possible variations with regard to magnetic disturbances.
